
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
```

# Logistic regression

## Dataset

```{r}
alc_data <- read.csv("./data/student/pormath.csv", stringsAsFactors = TRUE)
dim(alc_data)
colnames(alc_data)
```

## What explains high alcohol consumption
Graphs.
```{r message = FALSE, warning = FALSE}
# keep only variables of interest
alc <- alc_data %>% select(high_use,sex,famrel,absences,G3, goout, alc_use)
summary(select(alc_data,.data$G3,.data$absences,.data$famrel,.data$sex,.data$high_use,.data$goout))
gather(alc) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar()


# or
p_g3 <- ggplot(data=alc_data,aes(x=high_use,y=G3,fill=high_use))
p_g3 + geom_boxplot() + ylab("Final grade")
p_going_out <- ggplot(data=alc_data,aes(x=goout,fill=high_use))
p_going_out + geom_bar() + ylab("Going out")
p_studytime <- ggplot(data=alc_data,aes(x=studytime,fill=high_use))
p_studytime + geom_bar()
p_internet <- ggplot(data=alc_data,aes(x=internet,fill=high_use))
p_internet + geom_bar()
```

Crosstables
```{r}
table1 <- table(alc_data$sex, alc_data$high_use,dnn=c("sex","high_alc"))
addmargins(round(prop.table(table1)*100,1)) 

table2 <- table(alc_data$famrel, alc_data$high_use,dnn=c("famrel","high_alc"))
addmargins(round(prop.table(table2)*100,1)) 

table(alc_data$goout, alc_data$high_use)
```
Boxplots

The Box plots show the associations between the continuous variables grade and absences with alcohol consumption, drawn separately for females and males. 
```{r}
# initialize a plot of high_use and G3, grouping by sex
g1 <- ggplot(alc, aes(x = high_use, y = G3,col=sex))

# define the plot as a boxplot and draw it
g1 + geom_boxplot() + ylab("grade") +
ggtitle("Student grades by alcohol consumption and sex")
```

## Logistic regression model - what causes high alcohol consumption

```{r}
logistic <- glm(high_use ~  famrel + absences + G3 + goout + sex, data = alc, family = "binomial")

# print out a summary of the model
summary(logistic)

better_logistic <- glm(high_use ~  famrel + absences  + goout + sex, data = alc, family = "binomial")

```
Odd ratios
```{r}
# compute odds ratios (OR)
OR <- exp(coef(logistic))

# compute confidence intervals (CI)
CI <- logistic %>% confint %>% exp

# print out the odds ratios with their confidence intervals
cbind(OR, CI)

OR2 <- exp(coef(better_logistic))

# compute confidence intervals (CI)
CI2 <- better_logistic %>% confint %>% exp

# print out the odds ratios with their confidence intervals
cbind(OR2, CI2)

```
```{r}
#predict() the probability of high_use
probabilities <- predict(logistic, type = "response")

# add the predicted probabilities to 'alc'
alc <- mutate(alc, probability = probabilities)

# use the probabilities to make a prediction of high_use
alc <- mutate(alc, prediction = probability > 0.5)

# tabulate the target variable versus the predictions
table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table %>% addmargins()
```
I conduct a simple model validation by tabulating the actual outcomes versus model predictions. The proportion of incorrectly predicted students is around 20 percent (15.7 + 4.3). Thus, I would next search for additional explanatory variables!

Loss function:
```{r}
# define a loss function (mean prediction error)
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

# call loss_func to compute the average number of wrong predictions in the (training) data
loss_func(class = alc$high_use, prob = alc$probability)

```
Cross validation:
```{r}
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

# compute the average number of wrong predictions in the (training) data
loss_func(class = alc$high_use, prob = alc$probability)

# K-fold cross-validation
library(boot)
cv <- cv.glm(data = alc, cost = loss_func, glmfit = logistic, K = 10)

# average number of wrong predictions in the cross validation
cv$delta[1]
```

```{r}
alc2 <- alc_data %>% select(G3, goout,studytime,internet, high_use, absences, health, internet, nursery, romantic, sex)

logistic2 <- glm(high_use ~  internet + nursery + goout + sex, data = alc2, family = "binomial")
summary(logistic2)

#predict() the probability of high_use
probabilities2 <- predict(logistic2, type = "response")

# add the predicted probabilities to 'alc'
alc2 <- mutate(alc2, probability = probabilities2)

# use the probabilities to make a prediction of high_use
alc2 <- mutate(alc2, prediction = probability > 0.5)

loss_func(class = alc2$high_use, prob = alc2$probability)

cv2 <- cv.glm(data = alc2, cost = loss_func, glmfit = logistic2, K = 5)

# average number of wrong predictions in the cross validation
cv2$delta[1]

```
















